---
title: "Urbanisation weakens phenotypic selection but enhances the adaptive potential of fruit traits in a seed-dispersal mutualism"
subtitle: "Supplementary material"
author: 'Code author: Facundo X. Palacio'
date: "`r Sys.Date()`"
output: 
  tufte::tufte_html:
    toc: true
---

\ 

This document presents an overview of the analyses in the main text. It highlights the essential R code used to conduct these analyses, but does not display code related to table and plot aesthetics. The code for this document is provided in **Palacio&Ordano_urbanisation_Passiflora.Rmd** and is available at https://github.com/facuxpalacio/Passiflora_urbanisation. 


```{r setup, include = FALSE}
# Suppress messages, warnings and errors for document aesthetics.
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE, echo = TRUE)
```

```{r packages, results = 'hide'}
# Load libraries
library(dplyr) # Data management
library(corrr) # Computing correlations
library(ggraph) # Plotting graphs
library(igraph) # Plotting graphs
library(lme4) # Fitting mixed models
library(lmerTest) # Adding p-values in mixed models
library(multcomp) # Computing multiple comparisons
library(MASS) # Fitting negative binomial models
library(ggplot2) # Plotting
library(ggpubr) # Arranging multiple panels in the same plot
library(piecewiseSEM) # Structural equations
library(boot) # Bootstrapping
library(visreg) # Plotting fitness surface
```

## Correlations among fruit and seed traits

```{r corr}
# Load fruit data (fruit size, seed number)
fruits <- read.table("C:/RD/Palacio&Ordano_fruits_urbanisation.txt", head = T)

cors <- fruits %>% 
  filter(population == "urban") %>%
  select(mass_g, length, diam, seedn, pulp_sugar_bx, peel_sugar_bx) %>%
  correlate() %>% 
  stretch() %>%
  graph_from_data_frame(directed = FALSE)

ggraph(cors) +
  geom_edge_link(aes(edge_alpha = abs(r), edge_width = abs(r), color = r)) +
  guides(edge_alpha = "none", edge_width = "none") +
  scale_edge_colour_gradientn(limits = c(-1, 1), 
                              colors = c("firebrick2", "dodgerblue2")) +
  geom_node_point(color = "black", size = 8) +
  geom_node_text(aes(label = c("FM", "FL", "FD", "SN", "PuC", "PeC")), 
                 repel = TRUE)

cors <- fruits %>% 
  filter(population == "semiurban") %>%
  select(mass_g, length, diam, seedn, pulp_sugar_bx, peel_sugar_bx) %>%
  correlate() %>% 
  stretch() %>%
  graph_from_data_frame(directed = FALSE)

ggraph(cors) +
  geom_edge_link(aes(edge_alpha = abs(r), edge_width = abs(r), color = r)) +
  guides(edge_alpha = "none", edge_width = "none") +
  scale_edge_colour_gradientn(limits = c(-1, 1), 
                              colors = c("firebrick2", "dodgerblue2")) +
  geom_node_point(color = "black", size = 8) +
  geom_node_text(aes(label = c("FM", "FL", "FD", "SN", "PuC", "PeC")), 
                 repel = TRUE)

cors <- fruits %>% 
  filter(population == "rural") %>%
  select(mass_g, length, diam, seedn, pulp_sugar_bx, peel_sugar_bx) %>%
  correlate() %>% 
  stretch() %>%
  graph_from_data_frame(directed = FALSE)

ggraph(cors) +
  geom_edge_link(aes(edge_alpha = abs(r), edge_width = abs(r), color = r)) +
  guides(edge_alpha = "none", edge_width = "none") +
  scale_edge_colour_gradientn(limits = c(-1, 1), 
                              colors = c("firebrick2", "dodgerblue2")) +
  geom_node_point(color = "black", size = 8) +
  geom_node_text(aes(label = c("FM", "FL", "FD", "SN", "PuC", "PeC")), 
                 repel = TRUE)
```

\

# Influence of urbanisation on fruit and seed traits
```{r glmms}
# Load plant data (fruit crop size and removal)
plants <- read.table("C:/RD/Palacio&Ordano_plants_urbanisation.txt", head = TRUE)
plants$population <- as.factor(plants$population)

# Load fruit sugar data (it has information on raw concentrations of both fruit pulp and peel that needs to be preprocessed to obtain original concentrations)
sugar <- read.table("C:/RD/Palacio&Ordano_sugar_urbanisation.txt", head = T)
pulp <- subset(sugar, sample_type == "pulp")
peel <- subset(sugar, sample_type == "peel")

# Fruit crop size
mcrop <- glm.nb(crop ~ population, data = plants)
anova(mcrop)
multcrop <- glht(mcrop, linfct = mcp(population = "Tukey"))
summary(multcrop)

mdiam <- lmer(diam ~ population + (1|idplant), data = fruits)
anova(mdiam)
multdiam <- glht(mdiam, linfct = mcp(population = "Tukey"))
summary(multdiam)

mseed <- lmer(seedn ~ population + (1|idplant), data = fruits)
anova(mseed)

mbx_pulp <- lmer(sugar_bx_obs ~ population + (1|idplant), data = pulp)
anova(mbx_pulp)
multpulp <- glht(mbx_pulp, linfct = mcp(population = "Tukey"))
summary(multpulp)

mbx_peel <- lmer(true_sugar_bx ~ population + (1|idplant), data = peel)
anova(mbx_peel)
multpeel<- glht(mbx_peel, linfct = mcp(population = "Tukey"))
summary(multpeel)
```

\

## Boxplots of fruit and seed traits along with mixed model p-values

```{r boxplots}
crop <- ggplot(data = plants, aes(x = population, y = crop)) +
  geom_boxplot(color = "black", outlier.shape = NA, width = 0.5)+
  geom_jitter(position = position_jitter(0.2), color = "orange", 
              alpha = 0.6, size = 2) +
  xlab("Urbanisation level") + ylab("Fruit crop size") +
  geom_text(data = data.frame(x = 1:3, y = c(30, 130, 110)), 
            aes(x = x, y = y, label = c("a", "b", "b"))) + theme_bw()

diam <- ggplot(data = fruits, aes(x = population, y = diam)) +
  geom_boxplot(color = "black", outlier.shape = NA, width = 0.5) +
  geom_jitter(position = position_jitter(0.2), color = "orange", 
              alpha = 0.6, size = 2) +
  xlab("Urbanisation level") + ylab("Fruit diameter (mm)") +
  geom_text(data = data.frame(x = 1:3, y = c(44, 52, 47)), 
            aes(x = x, y = y, label = c("a", "b", "a"))) + theme_bw()

seedn <- ggplot(data = fruits, aes(x = population, y = seedn)) +
  geom_boxplot(color = "black", outlier.shape = NA, width = 0.5)+
  geom_jitter(position = position_jitter(0.2), color = "orange", 
              alpha = 0.6, size = 2) +
  xlab("Urbanisation level") + ylab("Seed number") +
  geom_text(data = data.frame(x = 1:3, y = c(315, 350, 360)), 
            aes(x = x, y = y, label = c("ns", "ns", "ns"))) + theme_bw()

pulp_bx <- ggplot(data = pulp, aes(x = population, y = sugar_bx_obs)) +
  geom_boxplot(color = "black", outlier.shape = NA, width = 0.5)+
  geom_jitter(position = position_jitter(0.2), color = "orange", 
              alpha = 0.6, size = 2) +
  xlab("Urbanisation level") + ylab("Pulp soluble carbohydrate content (?Bx)") +
  geom_text(data = data.frame(x = 1:3, y = c(22, 30, 29)), 
            aes(x = x, y = y, label = c("a", "b", "c"))) + theme_bw()

peel_bx <- ggplot(data = peel, aes(x = population, y = true_sugar_bx)) +
  geom_boxplot(color = "black", outlier.shape = NA, width = 0.5)+
  geom_jitter(position = position_jitter(0.2), color = "orange", 
              alpha = 0.6, size = 2) +
  xlab("Urbanisation level") + ylab("Peel soluble carbohydrate content (?Bx)") +
  geom_text(data = data.frame(x = 1:3, y = c(5, 8.7, 8)), 
            aes(x = x, y = y, label = c("a", "b", "b"))) + theme_bw()
  
ggarrange(crop, diam, seedn, pulp_bx, peel_bx, nrow = 3, ncol = 2)
```

\

## Relationship between urbanisation and bird-mediated selection using pieciewise structural equation models

```{r sems}
plants <- read.table("C:/RD/Palacio&Ordano_plants_urbanisation.txt", head = TRUE)
plants$population <- as.factor(plants$population)

plants <- na.omit(plants[, c(1, 4:6, 8:9, 11)])

### Raw estimates assuming different error distributions
# M1. Influence of urbanisation on bird visitation
M1 <- psem(
  # Influence of urbanisation on fruit traits
  glm.nb(crop ~ population, data = plants), 
  lm(x.diam ~ population, data = plants), 
  lm(x.nseeds ~ population, data = plants),
  lm(x.peel.sugar ~ population + crop, data = plants), 
  
  # Fruit traits and fitness
  glm(visits ~ population + crop + x.diam + x.nseeds + x.peel.sugar,
         family = poisson, data = plants),
  
  # Correlation between fitness measures
  fruit_rem %~~% visits,
  
  # Fruit trait correlations
  x.nseeds %~~% x.diam
)

resM1 <- summary(M1, direction = c("x.nseeds <- crop"))

# M2. Influence of urbanization on fruit removal
M2 <- psem(
  # Influence of urbanization on fruit traits
  glm.nb(crop ~ population, data = plants), 
  lm(x.diam ~ population, data = plants), 
  lm(x.nseeds ~ population, data = plants),
  lm(x.peel.sugar ~ population + crop, data = plants), 
  
  # Fruit traits and fitness
  glm(visits ~ population + crop + x.diam + x.nseeds + x.peel.sugar, 
      family = poisson, data = plants),
  
  # New paths: influence of fruit traits and urbanization on fruit removal
  glm.nb(fruit_rem ~ population + crop + x.diam + x.nseeds + x.peel.sugar, 
         data = plants),
  
  # Correlation between fitness measures
  fruit_rem %~~% visits,
  
  # Fruit trait correlations
  x.nseeds %~~% x.diam
)
  
resM2 <- summary(M2, direction = c("x.nseeds <- crop"))


# M3. Visits are influenced by fruit traits rather than by urbanisation,fruit removal is not influenced by urbanisation
M3 <- psem(
  # Influence of urbanisation on fruit traits
  glm.nb(crop ~ population, data = plants), 
  lm(x.diam ~ population, data = plants), 
  lm(x.nseeds ~ population, data = plants),
  lm(x.peel.sugar ~ population + crop, data = plants), 
  
  # Fruit traits and fitness
  glm(visits ~ crop + x.diam + x.nseeds + x.peel.sugar, 
      family = poisson, data = plants),
  
  # Correlation between fitness measures
  fruit_rem %~~% visits,
  
  # Fruit trait correlations
  x.nseeds %~~% x.diam)

resM3 <- summary(M3, direction = c("x.nseeds <- crop"))

# Model comparison
resM1$IC[2]
resM2$IC[2]
resM3$IC[2]

# Multigroup model (interaction)
M3multi <- psem(
    lm(x.peel.sugar ~ crop, data = plants), 
  
  # Fruit traits and fitness
  glm(visits ~ crop + x.diam + x.nseeds + x.peel.sugar, 
      family = poisson, data = plants),
  
  # Correlation between fitness measures
  fruit_rem %~~% visits,
  
  # Fruit trait correlations
  x.nseeds %~~% x.diam)

M3multigroup <- multigroup(M3multi, group = "population")
M3multigroup
```

\

### Estimate selection gradients from the structural equation
```{r sem gradients}
zdata <- na.omit(data.frame(population = plants$population,
                            zcrop = scale(plants$crop),
                            zdiam = scale(plants$x.diam),
                            znseeds = scale(plants$x.nseeds),
                            zpeel.sugar = scale(plants$x.peel.sugar),
                            wvisits = plants$visits/mean(plants$fruit_rem),
                            wfruit_rem = plants$fruit_rem/mean(plants$fruit_rem)))

M3 <- psem(
  # Influence of urbanization on fruit traits
  lm(zcrop ~ population, data = zdata), 
  lm(zdiam ~ population, data = zdata), 
  lm(znseeds ~ population, data = zdata),
  lm(zpeel.sugar ~ population + zcrop, data = zdata), 
  
  # Fruit traits and fitness
  lm(wvisits ~ zcrop + zdiam + znseeds + zpeel.sugar, data = zdata),
  
  # Correlation between fitness measures
  wfruit_rem %~~% wvisits,
  
  # Fruit trait correlations
  znseeds %~~% zdiam)

resM3 <- summary(M3)
resM3
```

\

## Natural selection analysis
### Opportunity for selection $I = \sigma^{2}_{w}$
```{r selection}
plants <- read.table("C:/RD/Palacio&Ordano_plants_urbanisation.txt", head = TRUE)
plants$population <- as.factor(plants$population)

plants %>% group_by(population) %>% summarise(Wfruit_removal = mean(fruit_rem),
                                              Wvisits = mean(visits),
                                              Ifruit_removal = var(fruit_rem/mean(fruit_rem)),
                                              Ivisits = var(visits/mean(visits)),
                                              n_plants = n(),
                                              sdIfruit_removal = Ifruit_removal/(2*n_plants),
                                              sdIvisits = Ivisits/(2*n_plants)) %>% 
  knitr::kable(., caption = "Opportunity for selection", digits = 2)
```

### Estimation of selection gradients
What follows is a function called `grad`, which computes selection gradients for a set of traits and populations based on bootstrap samples. For this function, the data must have the population identifier in the first column, the fitness in the second column, and the phenotypic traits in the remaining columns. The function returns a vector with linear, quadratic and correlational gradients, and represents the input for the function `boot` of the `boot` package. 
```{r gradients}
grad <- function(data, original = c(1:nrow(data))){
  data <- data[original, ] # for boot function
  traits  <- colnames(data)[-c(1:2)]
  z <- scale(data[, traits]) # standardise traits
  Wrel <- data[, 2]/mean(data[, 2]) # relative fitness
  df <- data.frame(z, Wrel)
  # linear gradients
  model.lin <- as.formula(paste("Wrel", paste(traits, collapse = " + "), sep = " ~ "))
  m1 <- lm(formula = model.lin, data = df)
  # nonlinear gradients
  part1 <- paste("(", paste(traits, collapse = " + "), ")^2", sep = "")
  part2 <- paste("I(0.5*(", traits, "^2))", sep = "", collapse = " + ")
  model.qua <- as.formula <- paste("Wrel", paste(part1, part2, sep = " + "), sep = " ~ ")
  m2 <- lm(formula = model.qua, data = df)
  
  sel.grad <- c(m1$coefficients[-1], m2$coefficients[-c(1:ncol(df))])
  return(sel.grad)
}

# Select traits
df <- plants[, c("population", "fruit_rem", "crop", "x.diam", "x.nseeds", "x.peel.sugar")]
idpop <- unique(df[, 1])
pops <- split(df, df[, 1])
npops <- length(pops)
sel.grad.pops <- list()

# Compute selection gradients for each population
for(i in 1:npops){
  data_pop <- subset(df, df[, 1] == idpop[i])
  sel.grad.pops[[i]] <- grad(data = data_pop)
}

population <- unique(plants$population)
names(sel.grad.pops) <- population
as.data.frame(sel.grad.pops) %>% knitr::kable(., digits = 2, caption = "Selection gradients")

# Confidence intervals for selection gradients based on bootstrap samples
set.seed(999)
nsamples <- 1000
boot.grad <- list()

for(i in 1:npops){
  data_pop <- subset(df, df[, 1] == idpop[i])
  boot.replicates <- boot(data = data_pop, statistic = grad, R = nsamples)
  
  # Confidence intervals
  CI <- list()
  ngrad <- length(boot.replicates$t0)
  for(j in 1:ngrad){
    CI[[j]] <- boot.ci(boot.replicates, conf = 0.95, type = "bca", index = j)$bca[4:5]
  }
 
  names(CI) <- names(boot.replicates$t0)
  
  # Table of gradients, standard errors, confidence intervals and p-values
  boot.grad[[i]] <- data.frame(gradient = boot.replicates$t0, 
                               se = apply(boot.replicates$t, 2, sd, na.rm = TRUE),
                               low.CI = as.numeric(t(as.data.frame(CI)[1, ])),
                               upp.CI = as.numeric(t(as.data.frame(CI)[2, ])),
                               p = apply(boot.replicates$t, 2, function(x) ifelse(mean(x, na.rm = T)>0, length(x[x<0])/nsamples, length(x[x>0])/nsamples)))
}

names(boot.grad) <- idpop
boot.grad %>% knitr::kable(., digits = 2, Caption = "Selection gradients and uncertainty based on bootstrap samples")

# Pepare gradients for plotting
boot_results <- do.call("rbind", boot.grad)
ngrad <- nrow(boot.grad[[1]])
boot_results$population <- sub("\\..*", "", rownames(boot_results))
boot_results$grad_type <- rep(c(rep("linear", 4), rep("nonlinear", 10)), npops)
traits  <- colnames(df)[-c(1:2)]
boot_results$trait <- rep(c(rep(traits, 2), 
                            paste(traits[1], traits[2]),
                            paste(traits[1], traits[3]),
                            paste(traits[1], traits[4]),
                            paste(traits[2], traits[3]),
                            paste(traits[2], traits[4]),
                            paste(traits[3], traits[4])), npops)

# Plot linear gradients
linear_results <- subset(boot_results, grad_type == "linear")
ggplot(data = linear_results, aes(x = population, y = gradient, col = trait)) +
  geom_point(position = position_dodge(0.5), size = 3) +
  geom_errorbar(aes(ymin = low.CI, ymax = upp.CI), width = 0, position = position_dodge(0.5)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Population") + ylab("Linear gradient") + theme_bw()

# Plot nonlinear gradients
nonlinear_results <- subset(boot_results, grad_type == "nonlinear")
ggplot(data = nonlinear_results, aes(x = population, y = gradient, col = trait)) +
  geom_point(position = position_dodge(0.5), size = 3) +
  geom_errorbar(aes(ymin = low.CI, ymax = upp.CI), 
                width = 0, position = position_dodge(0.5)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Population") + ylab("Nonlinear gradient") + ylim(-3, 3) + theme_bw()

# Non-linear selection on fruit crop size in the semi-urban population
semiu <- subset(plants, population == "semiurban")
semiu$zcrop <- as.numeric(scale(semiu$crop))
semiu$zdiam <- as.numeric(scale(semiu$x.diam))
semiu$zseeds <- as.numeric(scale(semiu$x.nseeds))
semiu$zpeel <- as.numeric(scale(semiu$x.peel.sugar))
semiu$wrel <- semiu$fruit_rem/mean(semiu$fruit_rem, na.rm=T)
nonlin.grad <- lm(wrel ~ zcrop + I(0.5*zcrop^2) + zdiam + I(0.5*zdiam^2) + zseeds + I(0.5*zseeds^2) + zpeel + I(0.5*zpeel^2) + zcrop:zdiam + zcrop:zseeds + zcrop:zpeel + zdiam:zseeds + zdiam:zpeel + zseeds:zpeel, data = semiu)

df <- data.frame(zcrop = seq(-1, 4, length = 500), zdiam = 0, zseeds = 0, zpeel = 0)
pred_fitness <- data.frame(zcrop = seq(-1, 4, length = 500), 
                           wrel = predict(quad.grad, newdata = df))

ggplot() + 
geom_point(data = semiu, aes(x = zcrop, y = wrel), size = 5) +
geom_line(data = pred_fitness, aes(x = zcrop, y = wrel), col = "blue", lwd = 1.5) +
xlab("Standardised fruit crop size") + 
ylab("Relative fitness (number of pecked fruits)") +
theme_bw()

visreg2d(nonlin.grad, xvar = "zcrop", yvar = "zpeel", type = "conditional", cond = list(zdiam = 0, zseeds = 0, zpeel = 0), xlab = "Standardised fruit crop size", main = "Semi-urban population", ylab = "Standardised peel soluble carbohydrate concentration")
points(semiu$zcrop, semiu$zpeel, pch = 19, cex = 2)
```

