---
title: "Urbanisation shapes phenotypic selection of fruit traits in a seed-dispersal mutualism"
subtitle: "Supplementary material"
author: 'Code author: Facundo X. Palacio'
date: "`r Sys.Date()`"
output: 
  tufte::tufte_html:
    toc: true
---

\ 

This document presents an overview of the analyses in the main text. It highlights the essential R code used to conduct these analyses, but does not display code related to table and plot aesthetics. The code for this document is provided in **Palacio&Ordano_urbanisation_Passiflora.Rmd** and is available at https://github.com/facuxpalacio/Passiflora_urbanisation. 


```{r setup, include = FALSE}
# Suppress messages, warnings and errors for document aesthetics.
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE, echo = TRUE)
```

```{r packages, results = 'hide'}
# Load libraries
library(plyr) # Data management
library(dplyr) # Data management
library(tidyr) # Data management
library(corrr) # Computing correlations
library(ggraph) # Plots
library(igraph) # Plots
library(lme4) # Fitting mixed models
library(lmerTest) # Adding p-values in mixed models
library(multcomp) # Computing multiple comparisons
library(MASS) # Fitting negative binomial models
library(ggplot2) # Plotting
library(ggpubr) # Arranging multiple panels in the same plot
library(MVN) # Multivariate normality tests
library(piecewiseSEM) # Structural equations
library(fields) # Plotting fitness surfaces
library(mgcv) # Fitness landscapes
library(boot) # Bootstrapping
library(vegan) # Permanova
```

\

## Correlations among fruit and seed traits

```{r corr_urb, fig.cap = "Correlation network between fruit and seed traits in an urban *Passiflora caerulea* population. FM: fruit mass, FL: fruit length, FD: fruit diameter, SN: seed number, PuC: pulp soluble carbohydrate concentration, PeC: peel soluble carbohydrate concentration"}
# Load fruit data (fruit size, seed number)
fruits <- read.table("data/Palacio&Ordano_fruits_urbanisation.txt", head = T)

# Urban population
cors <- fruits %>% 
  filter(population == "urban") %>%
  dplyr::select(mass_g, length, diam, seedn, pulp_sugar_bx, peel_sugar_bx) %>%
  correlate() %>% 
  stretch() %>%
  graph_from_data_frame(directed = FALSE)

ggraph(cors) +
  geom_edge_link(aes(edge_alpha = abs(r), edge_width = abs(r), color = r)) +
  guides(edge_alpha = "none", edge_width = "none") +
  scale_edge_colour_gradientn(limits = c(-1, 1), 
                              colors = c("firebrick2", "dodgerblue2")) +
  geom_node_point(color = "black", size = 8) +
  geom_node_text(aes(label = c("FM", "FL", "FD", "SN", "PuC", "PeC")), 
                 repel = TRUE) + ggtitle("Urban population")
```

```{r corr_semiu, fig.cap = "Correlation network between fruit and seed traits in a semiurban *Passiflora caerulea* population. FM: fruit mass, FL: fruit length, FD: fruit diameter, SN: seed number, PuC: pulp soluble carbohydrate concentration, PeC: peel soluble carbohydrate concentration"}
# Semiurban population
cors <- fruits %>% 
  filter(population == "semiurban") %>%
  dplyr::select(mass_g, length, diam, seedn, pulp_sugar_bx, peel_sugar_bx) %>%
  correlate() %>% 
  stretch() %>%
  graph_from_data_frame(directed = FALSE)

ggraph(cors) +
  geom_edge_link(aes(edge_alpha = abs(r), edge_width = abs(r), color = r)) +
  guides(edge_alpha = "none", edge_width = "none") +
  scale_edge_colour_gradientn(limits = c(-1, 1), 
                              colors = c("firebrick2", "dodgerblue2")) +
  geom_node_point(color = "black", size = 8) +
  geom_node_text(aes(label = c("FM", "FL", "FD", "SN", "PuC", "PeC")), 
                 repel = TRUE) + ggtitle("Semi-urban population")
```

```{r corr_rur, fig.cap = "Correlation network between fruit and seed traits in a rural *Passiflora caerulea* population. FM: fruit mass, FL: fruit length, FD: fruit diameter, SN: seed number, PuC: pulp soluble carbohydrate concentration, PeC: peel soluble carbohydrate concentration"}
# Rural population
cors <- fruits %>% 
  filter(population == "rural") %>%
  dplyr::select(mass_g, length, diam, seedn, pulp_sugar_bx, peel_sugar_bx) %>%
  correlate() %>% 
  stretch() %>%
  graph_from_data_frame(directed = FALSE)

ggraph(cors) +
  geom_edge_link(aes(edge_alpha = abs(r), edge_width = abs(r), color = r)) +
  guides(edge_alpha = "none", edge_width = "none") +
  scale_edge_colour_gradientn(limits = c(-1, 1), 
                              colors = c("firebrick2", "dodgerblue2")) +
  geom_node_point(color = "black", size = 8) +
  geom_node_text(aes(label = c("FM", "FL", "FD", "SN", "PuC", "PeC")), 
                 repel = TRUE) + ggtitle("Rural population")
```

\

# Influence of urbanisation on fruit and seed traits
```{r glmms}
# Load plant data (fruit crop size and removal)
plants <- read.table("data/Palacio&Ordano_plants_urbanisation.txt", head = TRUE)
plants$population <- as.factor(plants$population)

# Load fruit sugar data (it has information on raw concentrations of both fruit pulp and peel that needs to be preprocessed to obtain original concentrations)
sugar <- read.table("data/Palacio&Ordano_sugar_urbanisation.txt", head = T)
pulp <- subset(sugar, sample_type == "pulp")
peel <- subset(sugar, sample_type == "peel")

# Fruit crop size
mcrop <- glm.nb(crop ~ population, data = plants)
anova(mcrop)
multcrop <- glht(mcrop, linfct = mcp(population = "Tukey"))
summary(multcrop)

# Fruit size
mdiam <- lmer(diam ~ population + (1|idplant), data = fruits)
anova(mdiam)
multdiam <- glht(mdiam, linfct = mcp(population = "Tukey"))
summary(multdiam)

# Seed number
mseed <- lmer(seedn ~ population + (1|idplant), data = fruits)
anova(mseed)

# Sugar concentration
mbx_pulp <- lmer(sugar_bx_obs ~ population + (1|idplant), data = pulp)
anova(mbx_pulp)
multpulp <- glht(mbx_pulp, linfct = mcp(population = "Tukey"))
summary(multpulp)

mbx_peel <- lmer(true_sugar_bx ~ population + (1|idplant), data = peel)
anova(mbx_peel)
multpeel<- glht(mbx_peel, linfct = mcp(population = "Tukey"))
summary(multpeel)

# Plant size
mlength <- lm(length ~ population, data = plants)
anova(mlength)
multlength <- glht(mlength, linfct = mcp(population = "Tukey"))
summary(multlength)
```

\

## Boxplots of fruit and seed traits along with mixed model p-values
```{r boxplots}
crop <- ggplot(data = plants, aes(x = population, y = crop)) +
  geom_jitter(position = position_jitter(0.2), color = "orange", 
              alpha = 0.6, size = 2) +
  geom_boxplot(color = "black", outlier.shape = NA, width = 0.5, alpha = 0.3) +
  xlab("Population") + ylab("Fruit crop size") +
  geom_text(data = data.frame(x = 1:3, y = c(30, 130, 110)), 
            aes(x = x, y = y, label = c("a", "b", "b"))) + theme_bw()

length <- ggplot(data = plants, aes(x = population, y = length)) +
  geom_jitter(position = position_jitter(0.2), color = "orange", 
              alpha = 0.6, size = 2) +
  geom_boxplot(color = "black", outlier.shape = NA, width = 0.5, alpha = 0.3) +
  xlab("Population") + ylab("Plant length (m)") +
  geom_text(data = data.frame(x = 1:3, y = c(7, 5.2, 10.2)), 
            aes(x = x, y = y, label = c("a", "a", "b"))) + theme_bw()

diam <- ggplot(data = fruits, aes(x = population, y = diam)) +
  geom_jitter(position = position_jitter(0.2), color = "orange", 
              alpha = 0.6, size = 2) +
  geom_boxplot(color = "black", outlier.shape = NA, width = 0.5, alpha = 0.3) +
  xlab("Population") + ylab("Fruit diameter (mm)") +
  geom_text(data = data.frame(x = 1:3, y = c(44, 52, 47)), 
            aes(x = x, y = y, label = c("a", "b", "a"))) + theme_bw()

seedn <- ggplot(data = fruits, aes(x = population, y = seedn)) +
  geom_jitter(position = position_jitter(0.2), color = "orange", 
              alpha = 0.6, size = 2) +
  geom_boxplot(color = "black", outlier.shape = NA, width = 0.5, alpha = 0.3) +
  xlab("Population") + ylab("Seed number") +
  geom_text(data = data.frame(x = 1:3, y = c(315, 350, 360)), 
            aes(x = x, y = y, label = c("ns", "ns", "ns"))) + theme_bw()

pulp_bx <- ggplot(data = pulp, aes(x = population, y = sugar_bx_obs)) +
  geom_jitter(position = position_jitter(0.2), color = "orange", 
              alpha = 0.6, size = 2) +
  geom_boxplot(color = "black", outlier.shape = NA, width = 0.5, alpha = 0.3) +
  xlab("Population") + ylab("Pulp soluble carbohydrate content (ºBx)") +
  geom_text(data = data.frame(x = 1:3, y = c(22, 30, 29)), 
            aes(x = x, y = y, label = c("a", "b", "c"))) + theme_bw()

peel_bx <- ggplot(data = peel, aes(x = population, y = true_sugar_bx)) +
  geom_jitter(position = position_jitter(0.2), color = "orange", 
              alpha = 0.6, size = 2) +
  geom_boxplot(color = "black", outlier.shape = NA, width = 0.5, alpha = 0.3) +
  xlab("Population") + ylab("Peel soluble carbohydrate content (ºBx)") +
  geom_text(data = data.frame(x = 1:3, y = c(5, 8.7, 8)), 
            aes(x = x, y = y, label = c("a", "b", "b"))) + theme_bw()
  
ggarrange(crop, length, diam, seedn, pulp_bx, peel_bx, nrow = 3, ncol = 3)
```

\

# Influence of urbanisation on bird visitation and fruit removal (Supplementary material)

```{r removal}
plants <- read.table("data/Palacio&Ordano_plants_urbanisation.txt", head = TRUE)
plants$population <- as.factor(plants$population)

mvisits <- glm(visits ~ population, family = poisson, data = plants)
anova(mvisits, test = "Chisq")
multvisits <- glht(mvisits, linfct = mcp(population = "Tukey"))
summary(multvisits)

mrem <- glm.nb(fruit_rem ~ population, data = plants)
anova(mrem)
multrem <- glht(mrem, linfct = mcp(population = "Tukey"))
summary(multrem)

visits <- ggplot(data = plants, aes(x = population, y = visits)) +
  geom_jitter(position = position_jitter(0.2), color = "orange", 
              alpha = 0.6, size = 2) +
  geom_boxplot(color = "black", outlier.shape = NA, width = 0.5, alpha = 0.3) +
  xlab("Population") + ylab("Number of visits") +
  geom_text(data = data.frame(x = 1:3, y = c(0.5, 1.4, 0.5)), 
            aes(x = x, y = y, label = c("a", "b", "ab"))) + theme_bw()

fruit_rem <- ggplot(data = plants, aes(x = population, y = fruit_rem)) +
  geom_jitter(position = position_jitter(0.2), color = "orange", 
              alpha = 0.6, size = 2) +
  geom_boxplot(color = "black", outlier.shape = NA, width = 0.5, alpha = 0.3) +
  xlab("Population") + ylab("Number of fruits pecked") +
  geom_text(data = data.frame(x = 1:3, y = c(15, 75, 28)), 
            aes(x = x, y = y, label = c("a", "b", "b"))) + theme_bw()

ggarrange(visits, fruit_rem, nrow = 1, ncol = 2)

```


\

## Relationship between urbanisation and bird-mediated selection using pieciewise structural equation models
```{r sems}
plants <- read.table("data/Palacio&Ordano_plants_urbanisation.txt", head = TRUE)
plants$population <- as.factor(plants$population)
plants <- na.omit(plants)

# M1. Visits are influenced by fruit traits rather than by urbanisation, fruit removal is not influenced by urbanisation
M1 <- psem(
  # Influence of urbanisation on fruit traits 
  glm.nb(crop ~ population + length, data = plants), 
  lm(length ~ population, data = plants),
  lm(x.diam ~ population, data = plants), 
  lm(x.nseeds ~ population, data = plants),
  lm(x.peel.sugar ~ population + crop, data = plants), 

  # Influence of fruit traits on bird visitation
  glm(visits ~ crop + x.diam + x.nseeds + x.peel.sugar, family = poisson, data = plants),
  
  # Fruit traits, bird visitation and fitness
  glm.nb(fruit_rem ~ crop + x.diam + x.nseeds + x.peel.sugar + length + visits, data = plants),
  
  # Fruit trait correlations
  x.nseeds %~~% x.diam)

resM1 <- summary(M1, direction = c("x.nseeds <- crop"))

# M2. Influence of urbanisation on fruit removal
M2 <- psem(
  # Influence of urbanisation on fruit traits and bird visitation
  glm.nb(crop ~ population + length, data = plants), 
  lm(length ~ population, data = plants),
  lm(x.diam ~ population, data = plants), 
  lm(x.nseeds ~ population, data = plants),
  lm(x.peel.sugar ~ population + crop, data = plants),
 
  # Urbanisation and bird visitation 
  glm(visits ~ crop + x.diam + x.nseeds + x.peel.sugar + population, family = poisson, data = plants),
  
  # New path: influence of urbanisation on fitness
  glm.nb(fruit_rem ~ crop + x.diam + x.nseeds + x.peel.sugar + length + visits + population, data = plants),
  
  # Fruit trait correlations
  x.nseeds %~~% x.diam)
  
resM2 <- summary(M2, direction = c("x.nseeds <- crop"))

# M3. Influence of urbanisation on bird visitation
M3 <- psem(
  # Influence of urbanisation on fruit traits and bird visitation
  glm.nb(crop ~ population + length, data = plants),
  lm(length ~ population, data = plants),
  lm(x.diam ~ population, data = plants), 
  lm(x.nseeds ~ population, data = plants),
  lm(x.peel.sugar ~ population + crop, data = plants),
 
  # New path: urbanisation and bird visitation 
  glm(visits ~ population, family = poisson, data = plants),
  
  # Fruit traits, bird visitation and fitness
  glm.nb(fruit_rem ~ crop + x.diam + x.nseeds + x.peel.sugar + length + visits, data = plants),
  
  # Fruit trait correlations
  x.nseeds %~~% x.diam)

resM3 <- summary(M3, direction = c("x.nseeds <- crop"))

# M4. Urbanisation influences fruit traits, bird visitation and fruit removal (combination of M2 and M3)
M4 <- psem(
  # Influence of urbanisation on fruit traits and bird visitation
  glm.nb(crop ~ population + length, data = plants), 
  lm(length ~ population, data = plants),
  lm(x.diam ~ population, data = plants), 
  lm(x.nseeds ~ population, data = plants),
  lm(x.peel.sugar ~ population + crop, data = plants),
 
  # Urbanisation and bird visitation 
  glm(visits ~ population, family = poisson, data = plants),
  
  # Influence of urbanisation on fitness
  glm.nb(fruit_rem ~ crop + x.diam + x.nseeds + x.peel.sugar + length + visits + population, data = plants),
  
  # Fruit trait correlations
  x.nseeds %~~% x.diam)

resM4 <- summary(M4, direction = c("x.nseeds <- crop"))

# Model comparison based on AICc
resM1$IC[2]
resM2$IC[2]
resM3$IC[2]
resM4$IC[2]

resM4 

# Multigroup model (interaction)
M4multi <- psem(
  # Influence of urbanisation on fruit traits and bird visitation
  glm.nb(crop ~ length, data = plants), 
  lm(x.peel.sugar ~ crop, data = plants),
   
  # Influence of urbanisation on fitness
  glm.nb(fruit_rem ~ crop + x.diam + x.nseeds + x.peel.sugar + length + visits, data = plants),
  
  # Fruit trait correlations
  x.nseeds %~~% x.diam)

M4multigroup <- multigroup(M4multi, group = "population")
M4multigroup

# Standardised estimates
plants <- plants %>% mutate(z.diam = scale(x.diam)[,1]) %>% mutate(z.nseeds = scale(x.nseeds)[,1]) %>%
mutate(z.peel.sugar = scale(x.peel.sugar)[,1]) %>%
mutate(z.crop = scale(crop)[,1]) %>%
mutate(z.length = scale(length)[,1]) %>%
mutate(z.fruit_rem = scale(fruit_rem)[,1]) %>%
mutate(z.visits = scale(visits)[,1])

# One model per population
lp <- subset(plants, population == "urban")
ra <- subset(plants, population == "semiurban")
mg <- subset(plants, population == "rural")

zM4_lp <- psem(
  # Influence of urbanisation on fruit traits and bird visitation
  lm(z.crop ~ z.length, data = lp), 
  lm(z.peel.sugar ~ z.crop, data = lp),
  
  # Influence of urbanisation on fitness
  lm(z.fruit_rem ~ z.crop + z.diam + z.nseeds + z.peel.sugar + z.length + z.visits, data = lp),
  
  # Fruit trait correlations
  z.nseeds %~~% z.diam)

reszM4_lp <- summary(zM4_lp)
reszM4_lp

zM4_ra <- psem(
  # Influence of urbanisation on fruit traits and bird visitation
  lm(z.crop ~ z.length, data = ra), 
  lm(z.peel.sugar ~ z.crop, data = ra),
  
  # Influence of urbanisation on fitness
  lm(z.fruit_rem ~ z.crop + z.diam + z.nseeds + z.peel.sugar + z.length + z.visits, data = ra),
  
  # Fruit trait correlations
  z.nseeds %~~% z.diam)

reszM4_ra <- summary(zM4_ra)

zM4_mg <- psem(
  # Influence of urbanisation on fruit traits and bird visitation
  lm(z.crop ~ z.length, data = mg), 
  lm(z.peel.sugar ~ z.crop, data = mg),
  
  # Influence of urbanisation on fitness
  lm(z.fruit_rem ~ z.crop + z.diam + z.nseeds + z.peel.sugar + z.length + z.visits, data = mg),
  
  # Fruit trait correlations
  z.nseeds %~~% z.diam)

reszM4_mg <- summary(zM4_mg)
reszM4_mg

# Relationship between fruit crop and removal for each population
population <- c("Urban", "Semi-urban", "Rural")
coeffs <- rbind(reszM4_lp$coeff[3, ], reszM4_ra$coeff[3, ], reszM4_mg$coeff[3, ])
data.frame(population, coeffs)
```

\

## Natural selection analysis
### Opportunity for selection
```{r selection}
plants <- read.table("data/Palacio&Ordano_plants_urbanisation.txt", head = TRUE)
plants$population <- as.factor(plants$population)

plants %>% group_by(population) %>% summarise(Wfruit_removal = mean(fruit_rem), 
                                              Ifruit_removal = var(fruit_rem/mean(fruit_rem)),
                                              n_plants = dplyr::n(),
                                              sdIfruit_removal = Ifruit_removal/(2*n_plants)) %>% 
  knitr::kable(., caption = "Opportunity for selection", digits = 2)

plants <- plants %>% group_by(population) %>% mutate(wrel = fruit_rem/mean(fruit_rem))
bartlett.test(wrel ~ population, data = plants)
```

### Step 1. Relative fitness and standardisation of phenotypic traits (mean = 0, variance = 1)
```{r traits_fitness}
urban <- subset(plants, population == "urban")
semiurban <- subset(plants, population == "semiurban")
rural <- subset(plants, population == "rural")

# Urban
urban$W <- urban$fruit_rem
urban$wrel <- urban$W/mean(urban$W)

urban$x1 <- log(urban$crop)
urban$x2 <- urban$x.diam
urban$x3 <- urban$x.nseeds
urban$x4 <- urban$x.peel.sugar

urban$z1 <- (urban$x1 - mean(urban$x1, na.rm = T))/sd(urban$x1, na.rm = T)
urban$z2 <- (urban$x2 - mean(urban$x2, na.rm = T))/sd(urban$x2, na.rm = T)
urban$z3 <- (urban$x3 - mean(urban$x3, na.rm = T))/sd(urban$x3, na.rm = T)
urban$z4 <- (urban$x4 - mean(urban$x4, na.rm = T))/sd(urban$x4, na.rm = T)

# Semiurban
semiurban$W <- semiurban$fruit_rem
semiurban$wrel <- semiurban$W/mean(semiurban$W)

semiurban$x1 <- log(semiurban$crop)
semiurban$x2 <- semiurban$x.diam
semiurban$x3 <- semiurban$x.nseeds
semiurban$x4<- semiurban$x.peel.sugar

semiurban$z1 <- (semiurban$x1 - mean(semiurban$x1, na.rm =T))/sd(semiurban$x1, na.rm = T)
semiurban$z2 <- (semiurban$x2 - mean(semiurban$x2, na.rm =T))/sd(semiurban$x2, na.rm = T)
semiurban$z3 <- (semiurban$x3 - mean(semiurban$x3, na.rm =T))/sd(semiurban$x3, na.rm = T)
semiurban$z4 <- (semiurban$x4 - mean(semiurban$x4, na.rm =T))/sd(semiurban$x4, na.rm = T)

# Rural
rural$W <- rural$fruit_rem
rural$wrel <- rural$W/mean(rural$W)

rural$x1 <- log(rural$crop)
rural$x2 <- rural$x.diam
rural$x3 <- rural$x.nseeds
rural$x4 <- rural$x.peel.sugar

rural$z1 <- (rural$x1 - mean(rural$x1, na.rm = T))/sd(rural$x1, na.rm = T)
rural$z2 <- (rural$x2 - mean(rural$x2, na.rm = T))/sd(rural$x2, na.rm =T )
rural$z3 <- (rural$x3 - mean(rural$x3, na.rm = T))/sd(rural$x3, na.rm = T)
rural$z4 <- (rural$x4 - mean(rural$x4, na.rm = T))/sd(rural$x4, na.rm = T)
```

### Step 2. Testing multivariate normality of phenotypic traits
```{r mvn}
### Urban
ggplot(data = urban, aes(x = z1)) + xlim(-2, 4) +
       geom_histogram(aes(y = ..density..), col = "black", fill = "gray", binwidth = 0.5, alpha = 0.5) +  
       geom_density(col = "red", fill = "red", alpha = 0.3) + theme_classic()

ggplot(data = urban, aes(x = z2)) + xlim(-4, 4) +
  geom_histogram(aes(y = ..density..), col = "black", fill = "gray", binwidth = 0.5, alpha = 0.5) +  
  geom_density(col = "red", fill = "red", alpha = 0.3) + theme_classic()

ggplot(data = urban, aes(x = z3)) + xlim(-4, 4) +
  geom_histogram(aes(y = ..density..), col = "black", fill = "gray", binwidth = 0.5, alpha = 0.5) +  
  geom_density(col = "red", fill = "red", alpha = 0.3) + theme_classic()

ggplot(data = urban, aes(x = z4)) + xlim(-4, 4) +
  geom_histogram(aes(y = ..density..), col = "black", fill = "gray", binwidth = 0.5, alpha = 0.5) +  
  geom_density(col = "red", fill = "red", alpha = 0.3) + theme_classic()

  # Multivariate normality test
MVN::mvn(data.frame(urban$z1, urban$z2, urban$z3, urban$z4), mvnTest = "hz")$multivariateNormality

  # Q-Q plots
layout(matrix(1:4, 2, 2))
qqnorm(urban$z1, cex = 1.5, pch = 19, col = rgb(red = 0, green = 0, blue = 0, alpha = 0.3), main = "Fruit crop size")
qqline(urban$z1, col = "red", lwd = 2)
qqnorm(urban$z2, cex = 1.5, pch = 19, col = rgb(red = 0, green = 0, blue = 0, alpha = 0.3), main = "Mean fruit diameter")
qqline(urban$z2, col = "red", lwd = 2)
qqnorm(urban$z3, cex = 1.5, pch = 19, col = rgb(red = 0, green = 0, blue = 0, alpha = 0.3), main = "Mean seed number")
qqline(urban$z3, col = "red", lwd = 2)
qqnorm(urban$z4, cex = 1.5, pch = 19, col = rgb(red = 0, green = 0, blue = 0, alpha = 0.3), main = "Mean peel soluble carbohydrate concentration")
qqline(urban$z4, col = "red", lwd = 2)
layout(1)

### Semiurban
ggplot(data = semiurban, aes(x = z1)) + xlim(-2, 4) +
       geom_histogram(aes(y = ..density..), col = "black", fill = "gray", binwidth = 0.5, alpha = 0.5) +  
       geom_density(col = "red", fill = "red", alpha = 0.3) + theme_classic()

ggplot(data = semiurban, aes(x = z2)) + xlim(-4, 4) +
  geom_histogram(aes(y = ..density..), col = "black", fill = "gray", binwidth = 0.5, alpha = 0.5) +  
  geom_density(col = "red", fill = "red", alpha = 0.3) + theme_classic()

ggplot(data = semiurban, aes(x = z3)) + xlim(-4, 4) +
  geom_histogram(aes(y = ..density..), col = "black", fill = "gray", binwidth = 0.5, alpha = 0.5) +  
  geom_density(col = "red", fill = "red", alpha = 0.3) + theme_classic()

ggplot(data = semiurban, aes(x = z4)) + xlim(-4, 4) +
  geom_histogram(aes(y = ..density..), col = "black", fill = "gray", binwidth = 0.5, alpha = 0.5) +  
  geom_density(col = "red", fill = "red", alpha = 0.3) + theme_classic()

  # Multivariate normality test
MVN::mvn(data.frame(semiurban$z1, semiurban$z2, semiurban$z3, semiurban$z4), mvnTest = "hz")$multivariateNormality

  # Q-Q plots
layout(matrix(1:4, 2, 2))
qqnorm(semiurban$z1, cex = 1.5, pch = 19, col = rgb(red = 0, green = 0, blue = 0, alpha = 0.3), main = "Fruit crop size")
qqline(semiurban$z1, col = "red", lwd = 2)
qqnorm(semiurban$z2, cex = 1.5, pch = 19, col = rgb(red = 0, green = 0, blue = 0, alpha = 0.3), main = "Mean fruit diameter")
qqline(semiurban$z2, col = "red", lwd = 2)
qqnorm(semiurban$z3, cex = 1.5, pch = 19, col = rgb(red = 0, green = 0, blue = 0, alpha = 0.3), main = "Mean seed number")
qqline(semiurban$z3, col = "red", lwd = 2)
qqnorm(semiurban$z4, cex = 1.5, pch = 19, col = rgb(red = 0, green = 0, blue = 0, alpha = 0.3), main = "Mean peel soluble carbohydrate concentration")
qqline(semiurban$z4, col = "red", lwd = 2)
layout(1)

### Rural
ggplot(data = rural, aes(x = z1)) + xlim(-2, 4) +
       geom_histogram(aes(y = ..density..), col = "black", fill = "gray", binwidth = 0.5, alpha = 0.5) +  
       geom_density(col = "red", fill = "red", alpha = 0.3) + theme_classic()

ggplot(data = rural, aes(x = z2)) + xlim(-4, 4) +
  geom_histogram(aes(y = ..density..), col = "black", fill = "gray", binwidth = 0.5, alpha = 0.5) +  
  geom_density(col = "red", fill = "red", alpha = 0.3) + theme_classic()

ggplot(data = rural, aes(x = z3)) + xlim(-4, 4) +
  geom_histogram(aes(y = ..density..), col = "black", fill = "gray", binwidth = 0.5, alpha = 0.5) +  
  geom_density(col = "red", fill = "red", alpha = 0.3) + theme_classic()

ggplot(data = rural, aes(x = z4)) + xlim(-4, 4) +
  geom_histogram(aes(y = ..density..), col = "black", fill = "gray", binwidth = 0.5, alpha = 0.5) +  
  geom_density(col = "red", fill = "red", alpha = 0.3) + theme_classic()

  # Multivariate normality test
MVN::mvn(data.frame(rural$z1, rural$z2, rural$z3, rural$z4), mvnTest = "hz")$multivariateNormality

  # Q-Q plots
layout(matrix(1:4, 2, 2))
qqnorm(rural$z1, cex = 1.5, pch = 19, col = rgb(red = 0, green = 0, blue = 0, alpha = 0.3), main = "Fruit crop size")
qqline(rural$z1, col = "red", lwd = 2)
qqnorm(rural$z2, cex = 1.5, pch = 19, col = rgb(red = 0, green = 0, blue = 0, alpha = 0.3), main = "Mean fruit diameter")
qqline(rural$z2, col = "red", lwd = 2)
qqnorm(rural$z3, cex = 1.5, pch = 19, col = rgb(red = 0, green = 0, blue = 0, alpha = 0.3), main = "Mean seed number")
qqline(rural$z3, col = "red", lwd = 2)
qqnorm(rural$z4, cex = 1.5, pch = 19, col = rgb(red = 0, green = 0, blue = 0, alpha = 0.3), main = "Mean peel soluble carbohydrate concentration")
qqline(rural$z4, col = "red", lwd = 2)
layout(1)
```
\

### Step 3. Assessing collinearity of phenotypic traits
```{r collinearity}
  # Pearson correlation between traits
round(cor(cbind(urban$z1, urban$z2, urban$z3, urban$z4), use = "complete.obs"), 2)
round(cor(cbind(semiurban$z1, semiurban$z2, semiurban$z3, semiurban$z4), use = "complete.obs"), 2)
round(cor(cbind(rural$z1, rural$z2, rural$z3, rural$z4), use = "complete.obs"), 2)
```
\

### Step 4. Estimation of selection gradients, standard errors and p-values
What follows is the code from Morrissey & Goudie (2022) Evolution 76-7:1378–1390 to estimate gradients from log-linear models.
```{r gradients}
v <- function (m){m[lower.tri(m,diag=TRUE)]}

make.formula <- function (z, W = "W"){
  p<-length(z)
  iInd<-v(matrix(1:p,p,p,byrow=FALSE))
  jInd<-v(matrix(1:p,p,p,byrow=TRUE))
  terms<-array(dim=2*p+(p^2-p)/2)
  terms[1:p]<-z
    for (i in 1:(p+(p^2-p)/2)){
      if (iInd[i]==jInd[i]){
        terms[p+i]<-paste("I(0.5*",z[iInd[i]],"*",z[jInd[i]],")",sep="")
        } else {
          terms[p+i]<-paste("I(",z[iInd[i]],"*",z[jInd[i]],")",sep="")
        }
      }
  terms<-paste(terms,collapse="+")
  paste(W,"~",terms,sep="")
}

make.g<- function (gij,p,Sigma,warn=FALSE){
  g<-matrix(NA,p,p)
  g[lower.tri(g,diag=TRUE)]<-gij
  g[upper.tri(g)]<-t(g)[upper.tri(t(g))]
  if (warn){
    if (is.positive.definite(solve(Sigma)-g)==FALSE){
      print("Warning: Omega is not positive-definite.")
    }
    }
  g
}

# Function to estimate linear gradients
beta<- function (b,gij,mu,Sigma){
  p<-length(b)
  g<-make.g(gij=gij,p=p,Sigma=Sigma)
  tildeb<-b+g%*%mu
  Phi<-diag(p)-g%*%Sigma
  solve(Phi)%*%tildeb
}

# Function to estimate nonlinear gradients
gamma<- function (b,gij,mu,Sigma){
  p<-length(b)
  g<-make.g(gij=gij,p=p,Sigma=Sigma)
  tildeb<-b+g%*%mu
  Phi<-diag(p)-g%*%Sigma
  beta<-solve(Phi)%*%tildeb
  beta %*% t(beta) + solve(Phi)%*% g
}

# Function to compute gradients and their standard errors
grads<- function (b,gij,mu,Sigma,vcov,returnCov=FALSE){
  beta_hat<-beta(b,gij,mu,Sigma)
  gamma_hat<-gamma(b,gij,mu,Sigma)
  p<-length(b)
  s<-p+(p^2-p)/2
  phi<-diag(p)-make.g(gij,p)%*%Sigma
  Q<-solve(phi)
  iInd<-v(matrix(1:p,p,p,byrow=FALSE))
  jInd<-v(matrix(1:p,p,p,byrow=TRUE))
  J12<-matrix(0,p,s)
  
  for (m in 1:s){
  e<-matrix(0,s,1); e[m,1]<-1;
  C<-matrix(0,p,p);
  C[iInd[m],jInd[m]]<-1
  C[jInd[m],iInd[m]]<-1
  J12<-J12+Q%*%C%*%(Sigma%*%beta_hat+mu)%*%t(e)
  }
  J21<-matrix(0,s,p)
  
  for (k in 1:p){
  e<-matrix(0,p,1); e[k,1]<-1;
  J21<-J21+ v(beta_hat%*%t(Q[,k])+Q[,k]%*%t(beta_hat))%*%t(e)
  }
  J22<-matrix(0,s,s)
  
  for (m in 1:s){
    e<-matrix(0,s,1); e[m,1]<-1;
    C<-matrix(0,k,k);
  C[iInd[m],jInd[m]]<-1
  C[jInd[m],iInd[m]]<-1
  M<-Q%*%C%*%(Sigma%*%beta_hat+mu)%*%t(beta_hat)
  J22<-J22+v(M+t(M)+Q%*%C%*%t(Q))%*%t(e)
  }
  
  J<-matrix(NA,k+s,k+s)
  J[1:p,1:p]<-Q
  J[1:p,(p+1):(p+s)]<-J12
  J[(p+1):(p+s),1:p]<-J21
  J[(p+1):(p+s),(p+1):(p+s)]<-J22
  
  SEs<-sqrt(diag(J %*% vcov %*% t(J)))
  coef.names<-c(paste("beta",1:p),paste("gamma",iInd,jInd))
  res<-as.data.frame(cbind(coef.names,
  c(beta_hat,gamma_hat[lower.tri(gamma_hat,diag=TRUE)]),SEs))
  names(res)<-c("selection gradient","estimate","SE")
  res[,2]<-as.numeric(as.character(res[,2]))
  res[,3]<-as.numeric(as.character(res[,3]))
  if (returnCov==TRUE) res<-list(grads=res,vcov=J %*% vcov %*% t(J))
  return(res)
}

# Vector mean and variance-covariance matrix of traits
mu <- rep(0, 4)
Sigma_urb <- cov(na.omit(urban[, c("z1", "z2", "z3", "z4")]))
Sigma_semiu <- cov(na.omit(semiurban[, c("z1", "z2", "z3", "z4")]))
Sigma_rur <- cov(na.omit(rural[, c("z1", "z2", "z3", "z4")]))

# Poisson generalized linear models for each population
f <- make.formula(W = "W", z = c("z1", "z2", "z3", "z4"))
mod_urb <- glm(as.formula(f), data = urban, family = "quasipoisson")
mod_semiu <- glm(as.formula(f), data = semiurban, family = "quasipoisson")
mod_rur <- glm(as.formula(f), data = rural, family = "poisson")

# Coefficients of the GLM
b_urb <- coef(mod_urb)[2:5]
gij_urb <- coef(mod_urb)[6:15]

b_semiu <- coef(mod_semiu)[2:5]
gij_semiu <- coef(mod_semiu)[6:15]

b_rur <- coef(mod_rur)[2:5]
gij_rur <- coef(mod_rur)[6:15]

# Error variance-covariance matrix of the b and gij terms
sampVar_urb <- vcov(mod_urb)[2:15, 2:15]
sampVar_semiu <- vcov(mod_semiu)[2:15, 2:15]
sampVar_rur <- vcov(mod_rur)[2:15, 2:15]

# Selection gradients, standard errors and p-values
gradients_urb <- grads(b_urb, gij_urb, mu, Sigma_urb, sampVar_urb)
gradients_urb %>% mutate(p = summary(mod_urb)$coeff[-1, 4]) %>%
knitr::kable(., digits = 3, caption = "Selection gradients of the urban population")

gradients_semiu <- grads(b_semiu, gij_semiu, mu, Sigma_semiu, sampVar_semiu)
gradients_semiu %>% mutate(p = summary(mod_semiu)$coeff[-1, 4]) %>%
knitr::kable(., digits = 3, caption = "Selection gradients of the semiurban population")

gradients_rur <- grads(b_rur, gij_rur, mu, Sigma_rur, sampVar_rur)
gradients_rur %>% mutate(p = summary(mod_rur)$coeff[-1, 4]) %>%
knitr::kable(., digits = 3, caption = "Selection gradients of the rural population")

# Plot phenotype-fitness functions
### Urban population
#### Fruit crop size
new.z1 <- seq(min(urban$z1), max(urban$z1), length = 500)
pred.z1 <- predict(mod_urb, type = "response", newdata = data.frame(z1 = new.z1, z2 = 0, z3 = 0, z4 = 0), se = TRUE)
z1.low <- pred.z1$fit - 1.96*pred.z1$se
z1.upp <- pred.z1$fit + 1.96*pred.z1$se
plot(new.z1, pred.z1$fit, type = "n", xlab = "Standardised (log) fruit crop size", 
ylab = "Absolute fitness (number of fruits pecked)", ylim = c(0, 50))
polygon(x = c(new.z1, rev(new.z1)),
        y = c(z1.low, rev(z1.upp)),
        lty = 0, col = "#fff7ec")
points(urban$z1, urban$W, cex = 2.5, col = "gray70")
lines(new.z1, pred.z1$fit, lwd = 4,col = "darkorchid")

### Semiurban population
#### Fruit crop size
new.z1 <- seq(min(semiurban$z1), max(semiurban$z1), length = 500)
pred.z1 <- predict(mod_semiu, type = "response", newdata = data.frame(z1 = new.z1, z2 = 0, z3 = 0, z4 = 0), se = TRUE)
z1.low <- pred.z1$fit - 1.96*pred.z1$se
z1.upp <- pred.z1$fit + 1.96*pred.z1$se
plot(new.z1, pred.z1$fit, type = "n", xlab = "Standardised (log)fruit crop size", 
ylab = "Absolute fitness (number of fruits pecked)", ylim = c(0, 80))
polygon(x = c(new.z1, rev(new.z1)),
        y = c(z1.low, rev(z1.upp)),
        lty = 0, col = "#fff7ec")
points(semiurban$z1, semiurban$W, cex = 2.5, col = "gray70")
lines(new.z1, pred.z1$fit, lwd = 4,col = "darkorchid")

#### Mean fruit diameter x mean seed number
diamxseed <- data.frame(xdiam = semiurban$z2, xseed = semiurban$z3, W = semiurban$W)
surf <- cbind(diamxseed$xdiam, diamxseed$xseed)
m_diamxseed <- Tps(surf, diamxseed$W)
surface(m_diamxseed, xlab = "Standardised mean fruit diameter",
        ylab = "Standardised mean seed number", extrap = FALSE)
points(semiurban$z2, semiurban$z3, cex = 2)

### Rural population
#### Fruit crop size
new.z1 <- seq(-2, 2, length = 500)
pred.z1 <- predict(mod_rur, type = "response", newdata = data.frame(z1 = new.z1, z2 = 0, z3 = 0, z4 = 0), se = TRUE)
z1.low <- pred.z1$fit - 1.96*pred.z1$se
z1.upp <- pred.z1$fit + 1.96*pred.z1$se
plot(new.z1, pred.z1$fit, type = "n", xlab = "Standardised (log)fruit crop size", 
ylab = "Absolute fitness (number of fruits pecked)")
polygon(x = c(new.z1, rev(new.z1)),
        y = c(z1.low, rev(z1.upp)),
        lty = 0, col = "#fff7ec")
points(rural$z1, rural$W, cex = 2.5, col = "gray70")
lines(new.z1, pred.z1$fit, lwd = 4, col = "darkorchid")

#### Mean fruit diameter x mean seed number
diamxseed <- na.omit(data.frame(xdiam = rural$z2, xseed = rural$z3, W = rural$W))
surf <- cbind(diamxseed$xdiam, diamxseed$xseed)
m_diamxseed <- Tps(surf, diamxseed$W)
surface(m_diamxseed, xlab = "Standardised mean fruit diameter",
        ylab = "Standardised mean seed number", extrap = FALSE)
points(rural$z2, rural$z3, cex = 2)
```

\

## Similarity between bird assemblages

```{r sp_comp}
set.seed(1001)
birds <- read.table("data/Palacio&Ordano_birds_urbanisation.txt", head = TRUE)
birds$idpopulation <- as.factor(birds$idpopulation)

# Permanova
adonis2(birds[, 3:10] ~ idpopulation, data = birds, method = "bray", permutations = 1000) %>% knitr::kable(., digits = 3)

# Assumption of similar multivariate spread among populations
birds_dist <- vegdist(birds[, 3:10], method = "bray")
permutest(betadisper(birds_dist, birds$idpopulation), permutations = 1000, pairwise = TRUE) 

# Multiple comparisons and p-values adjustements
birds_urbsemiu <- subset(birds, idpopulation == "La_Plata"| idpopulation == "Ranchos")
birds_urbrur <- subset(birds, idpopulation == "La_Plata"| idpopulation == "Magdalena")
birds_semiurur <- subset(birds, idpopulation == "Ranchos"| idpopulation == "Magdalena")

urbsemiu <- adonis2(birds_urbsemiu[, 3:10] ~ idpopulation, data = birds_urbsemiu, method = "bray", permutations = 1000)
urbrur <- adonis2(birds_urbrur[, 3:10] ~ idpopulation, data = birds_urbrur, method = "bray", permutations = 1000)
semiurur <- adonis2(birds_semiurur[, 3:10] ~ idpopulation, data = birds_semiurur, method = "bray", permutations = 1000)

urbsemiu %>% knitr::kable(., digits = 3, caption = "Urban vs semiurban")
urbrur %>% knitr::kable(., digits = 3, caption = "Urban vs rural")
semiurur %>% knitr::kable(., digits = 3, caption = "Semiurban vs rural")

p_adonis <- c(urbsemiu[1, 5], urbrur[1, 5], semiurur[1, 5])
p.adjust(p_adonis, method = "holm")

# Rank abundance profiles
rural <- subset(birds, idpopulation == "Magdalena")[, -c(1,2)]
abu_rur <- sort(colSums(rural)/sum(rural), decreasing = TRUE)
semiurban <- subset(birds, idpopulation == "Ranchos")[, -c(1,2)]
abu_semiu <- sort(colSums(semiurban)/sum(semiurban), decreasing = TRUE)
urban <- subset(birds, idpopulation == "La_Plata")[, -c(1,2)]
abu_urb <- sort(colSums(urban)/sum(urban), decreasing = TRUE)
bird_abund <- data.frame(species = names(abu_rur), abu_rur, abu_semiu, abu_urb)

abu_rur
abu_semiu
abu_urb
```

